\chapter*{Acknowledgments}

I would like to express my sincere gratitude to all those who supported me during the development of this master’s thesis.

First and foremost, I am deeply grateful to my main advisor, Prof. Zhiling Lan, for her invaluable guidance throughout this work. She originally proposed the research idea and continuously provided direction, insightful feedback, and support. Her mentorship was fundamental in shaping both the technical content and the clarity of this thesis.

I also thank Prof. Michael Papka and Prof. Valerie Taylor for their valuable feedback. Prof. Papka offered insightful suggestions that significantly improved the quality and clarity of the visualizations, while Prof. Taylor provided early guidance that steered the research toward Large Language Model inference.

I am grateful to Prof. Davide Conficconi for his support and for sharing essential background and insights on hardware architectures, which contributed to a deeper understanding of the systems studied in this thesis.

My thanks also go to Dr. Varuni Sastry for her help in running experiments on limited-access accelerators and for her practical support during the experimental phase of this work.

I acknowledge Argonne National Laboratory for providing the necessary computational resources and an inspiring research environment. I also thank my colleagues for their collaboration and for the many stimulating discussions.

Finally, I am deeply thankful to my family and friends for their continuous support, encouragement, and patience throughout my academic journey.
\\ \\
GB
